---
title: "DSC1107 | FA3"
author: "Baybayon, Darlyn Antoinette"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Case Study: Major League Baseball

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggrepel)
  library(kableExtra)
  library(cowplot)
  library(readr)
  library(gridExtra)
})

```

#### Wrangle

##### 1.1 Import
- Import the data into a tibble called mlb_raw
- How many rows and columns does the data have?
- Does this match up with the data description given above?

**Solution**
```{r}
load("ml_pay.rdata")
mlb_raw <- as_tibble(ml_pay)
print(mlb_raw)

```

**There are 30 rows and 54 columns in the data which matches with the description.**

##### 1.2 Tidy

**Solution**
```{r}
mlb_aggregate <- mlb_raw %>%
  select(Team.name.2014, payroll, avgwin)%>% # select relevant columns
  rename( # rename for clarity
    team = Team.name.2014,
    payroll_aggregate = payroll,
    pct_wins_aggregate = avgwin
  )
mlb_aggregate


mlb_yearly <- mlb_raw %>%
  select(-payroll, -avgwin) %>% # remove irrelevant columns
  pivot_longer( # pivot to take all columns with years into a single column
    cols = matches("^p\\d{4}$|^X\\d{4}\\.pct$|^X\\d{4}$"), 
    names_to = "col_name", 
    values_to = "value", 
    values_drop_na = TRUE) %>%
  separate( # split the created column into three parts
    col = col_name, into = c("prefix", "year", "suffix"), 
    sep = "(?<=\\D)(?=\\d{4})|(?<=\\d{4})\\.", 
    fill = "right") %>%
  mutate( # create a new column which classifies the row based on the split columns
    col_class = case_when(
      prefix == "p" ~ "payroll",
      suffix == "pct" ~ "pct_wins",
      TRUE ~ "num_wins"
    )
  ) %>%
  select(team = Team.name.2014, year, col_class, value) %>% # select relevant columns
  pivot_wider( # pivot back to wider format and create separate columns based on the col_class
    names_from = col_class, values_from = value)
  
mlb_yearly
```


##### 1.3 Quality Control

•	Create a new tibble called mlb_aggregate_computed based on aggregating the data in mlb_yearly, containing columns named team, payroll_aggregate_computed, and pct_wins_aggregate_computed.

•	Ideally, mlb_aggregate_computed would match mlb_aggregate.  To check whether this is the case, join these two tibbles into mlb_aggregate_joined (which should have five columns: team, payroll_aggregate, pct_wins_aggregate, payroll_aggregate_computed, and pct_wins_aggregate_computed.)


**Solution**
```{r}
mlb_aggregate_computed <- mlb_yearly %>%
  group_by(team) %>%
  summarise(
    payroll_aggregate_computed = sum(payroll, na.rm = TRUE), # compute total payroll across 17 yrs
    pct_wins_aggregate_computed = mean(pct_wins, na.rm = TRUE) # compute mean pct wins across 17 yrs
  )

mlb_aggregate_joined <- mlb_aggregate %>% 
  left_join( # join the given aggregate and computed aggregate using left_join by col team
    mlb_aggregate_computed, by = 'team')

mlb_aggregate_joined

```

•	Create scatter plots of payroll_aggregate_computed versus payroll_aggregate and pct_wins_ aggregate_computed versus pct_wins_aggregate, including a 45◦ line in each. Display these scatter plots side by side, and comment on the relationship between the computed and provided aggregate statistics.

**Solution**
```{r}

payroll_plot <- ggplot( # plot payroll aggregate vs computed
  mlb_aggregate_joined, aes(x = payroll_aggregate * 1000, y = payroll_aggregate_computed)) +
  geom_point(color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + # add line with slope of 1
  labs(
    x = "Payroll Aggregate",
    y = "Payroll Aggregate Computed"
  )

pct_wins_plot <- ggplot( # plot pct wins aggregate vs computed
  mlb_aggregate_joined, aes(x = pct_wins_aggregate, y = pct_wins_aggregate_computed)) +
  geom_point(color = "blue") + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") + # add line with a slope of 1
  labs(
    x = "PCT Wins Aggregate",
    y = "PCT Wins Aggregate Computed"
  )

grid.arrange(payroll_plot, pct_wins_plot, ncol = 2)
```

**The scatter plots show that the given and computed payroll and win percentage are very similar as the points lie very closely along the straight diagonal line. There is slight discrepancy between the given and computed data likely due to data per year being rounded off. Computed payrolls tend to be higher than the given while discrepancy in pct wins is more random.**


#### 2 Explore

##### 2.1 Payroll across years
•	Plot payroll as a function of year for each of the 30 teams, faceting the plot by team and adding a red dashed horizontal line for the mean payroll across years of each team.

**Solution**
```{r, fig.width=9, fig.height=6}

mlb_yearly %>%
  group_by(team, year) %>%
  summarise( # calculate total payroll
    total_payroll = sum(payroll, na.rm = TRUE), .groups = "drop") %>%
  group_by(team) %>%
  mutate( # calculate mean payroll
    mean_payroll = mean(total_payroll, na.rm = TRUE)) %>%
  ggplot(aes(x = year, y = total_payroll, color = team)) +
  geom_line(aes(group = team), color = "blue") +
  facet_wrap(~team) +
  # add the horizontal red dashed line for the mean payroll
  geom_hline(aes(yintercept = mean_payroll), linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(
    title = "Payroll by Year for Each Team",
    x = "Year",
    y = "Payroll (in Millions)"
  )  +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))


```

•	Using dplyr, identify the three teams with the greatest payroll_aggregate_computed, and print a table of these teams and their payroll_aggregate_computed.

**Solution**
```{r}
mlb_aggregate_joined %>%
  select(team, payroll_aggregate_computed) %>% # select relevant columns
  arrange(desc(payroll_aggregate_computed)) %>% # arrange in desc order and display top 3
  head(3)

```

•		Using dplyr, identify the three teams with the greatest percentage increase in payroll from 1998 to 2014 (call it pct_increase), and print a table of these teams along with pct_increase as well as their payroll figures from 1998 and 2014.

**Solution**

```{r}
mlb_yearly %>% 
  select(team, year, payroll) %>% # select relevant columns
  pivot_wider( # pivot to wider format to see payroll from different years
     names_from = year, values_from = payroll, names_prefix = "payroll_")  %>%
  mutate( # create a new column for pct_increase
    pct_increase = (payroll_2014 - payroll_1998)/payroll_1998 * 100 
  ) %>%
  select(team, payroll_1998, payroll_2014, pct_increase)%>% # select relevant columns
  arrange(desc(pct_increase)) %>% # arrange in descending order and display top 3
  head(3)
```
•	How are the metrics payroll_aggregate_computed and pct_increase reflected in the plot above, and how can we see that the two sets of teams identified above are the top three in terms of these metrics?

**Teams with the highest total payroll . While teams with the highest percent increase in payroll would show a steeper increase in graphs.**


##### 2.2 Win percentage across years

•	Plot pct_wins as a function of year for each of the 30 teams, faceting the plot by team and adding a red dashed horizontal line for the average pct_wins across years of each team.

```{r, fig.width=9, fig.height=6}
mlb_yearly %>%
  select(team, year, pct_wins) %>% # select relevant columns
  group_by(team) %>% # group by team for next operations
  mutate( # create new column by taking mean of pct_wins across 17 yrs
    avg_pct_wins = mean(pct_wins, na.rm = TRUE)) %>%  
  ungroup() %>% # ungroup before plotting
  ggplot(aes(x = year, y = pct_wins, group = team)) + # plot yr vs pct wins
  geom_line(color = "blue") +  
  geom_hline(aes(yintercept = avg_pct_wins), linetype = "dashed", color = "red") + # add red line indicating mean pct wins
  facet_wrap(~ team) +  # facet by team
  labs(
    title = "Win Percentage Over Time by Team",
    x = "Year",
    y = "Win Percentage"
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust= 1)) # rotate x axis labels
```

•	Using dplyr, identify the three teams with the greatest pct_wins_aggregate_computed and print a table of these teams along with pct_wins_aggregate_computed.

**Solution**
```{r}
mlb_aggregate_joined %>% 
  select(team, pct_wins_aggregate_computed) %>% # select relevant cols
  arrange(desc(pct_wins_aggregate_computed)) %>% #arrange in desc order and display top 3
  head(3)
```

•	Using dplyr, identify the three teams with the most erratic pct_wins across years (as measured by the standard deviation, call it pct_wins_sd) and print a table of these teams along with pct_wins_sd.

**Solution**
```{r}
mlb_yearly %>%
  group_by(team) %>% # group by team to perform operations separately per team
  summarize(pct_wins_sd = sd(pct_wins, na.rm = TRUE)) %>% # calculate pct wins sd per team
  arrange(desc(pct_wins_sd)) %>% # arrange in desc order and display top 3
  head(3)

```

•	How are the metrics pct_wins_aggregate_computed and pct_wins_sd reflected in the plot above, and how can we see that the two sets of teams identified above are the top three in terms of these metrics?

**The teams with the greatest pct_wins_aggregate_computed have higher mean pct wins as observed in the red lines in the plot. The teams with the greatest pct_wins_sd notably have erratic graphs with many sharp fluctuations and more unpredictable behavior.**


##### 2.3 Win percentage vs. Payroll

•	Create a scatter plot of pct_wins versus payroll based on the aggregated data, labeling each point with the team name using geom_text_repel from the ggrepel package and adding the least squares line.


**Solution**

```{r, fig.width= 8, fig.height= 6}
ggplot(mlb_aggregate_computed, # plot payroll vs pct wins
       aes(x = payroll_aggregate_computed, y = pct_wins_aggregate_computed)) +
  geom_point(color = "blue", size = 2) +  
  geom_smooth(method = "lm", se = FALSE, color = "red", formula= y~x) + # add linear regression line
  geom_text_repel(aes(label = team), size =3, max.overlaps = 15) + # add labels per team
  labs(
    title = "Payroll vs. Win Percentage",
    x = "Payroll (in Billions)",
    y= "Win Percentage"
  )
```

•	Is the relationship between payroll and pct_wins positive or negative? Is this what you would expect, and why?

**The plot shows a positive relationship between payroll and pct_wins where win percentage increase as payroll increases. This is to be expected as teams with higher win percentage would have better players in their roster who are likely paid higher than other players.**


##### 2.4 Team Efficiency

•	Using dplyr, identify the three teams with the greatest efficiency, and print a table of these teams along with their efficiency, as well as their pct_wins_aggregate_computed and payroll_aggregate_computed.

**Solution**
```{r}

mlb_aggregate_computed %>%
  group_by(team) %>%
  mutate( # add a new col and compute team efficiency
    team_efficiency = pct_wins_aggregate_computed / payroll_aggregate_computed
  ) %>%
  arrange(desc(team_efficiency)) %>% # arrange in desc order and display top 3
  head(3)

```

**The three teams with top efficiency are Miami Marlins, Oakland Athletics, and Tampa Bay Rays**

•	In what sense do these three teams appear efficient in the previous plot?

**The pct_wins vs. payroll plot shows that these teams have higher win percentages compared to other teams with thin the same range of low payroll amount. Miami Marlins and Oakland Athletics are also plotted above the regression line, which implies that they are performing better than expected of their payrolls. Tampa Bay Rays falls a little short but they perform nearly as expected.**






